Informing policy via dynamic models: Eliminating cholera in Haiti

### 2022-09-15 ms.Rnw
Major structural changes

lee1_epi_mif <- bake(
  file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
  {
    foreach(
      i = 1:nrow(epi_params),
      .packages = c('pomp'),
      .combine = c
    ) %dopar% {
      r_params <- unlist(epi_params[i, ])
      coef(h1_epi) <- r_params
      mif2(
        h1_epi,
        Np = NP_MIF,
        Nmif = NMIF,
        cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
        rw.sd = epi_rw
      )
    }
  }
)

@@

lee_epi_ll <- bake(
  file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {

@@

    lee_epi_ll
  }
)

@@

lee1_end_mif <- bake(
  file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
  {
    foreach(
      i = 1:nrow(end_start),
      .packages = c('pomp'),
      .combine = c
    ) %dopar% {

@@

      coef(h1_end) <- r_params
      mif2(
        h1_end,
        Np = NP_MIF,
        Nmif = NMIF,
        cooling.type = "hyperbolic",
        cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
        rw.sd = end_rw
      )
      # output[['m2']] <- m2
      #
      # coef(h1_end) <- coef(m2)
      # sims <- simulate(  # Step 5
      #   h1_end, nsim = N_LEE_SIMS, format = 'data.frame',
      #   params = coef(m2)
      # )
      # sims$param_set <- i
      #
      # output[['sims']] <- sims
      # output
    }
  }
)

@@

lee_end_ll <- bake(
  file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {

@@

    lee_end_ll
  }
)

@@

We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models.
In Sec.~\ref{sec:results}, we use improved model fits to project cholera incidence in Haiti under various vaccination campaigns.
We then conclude with a discussion on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:discussion}.

@@

Proposed mechanistic structures form a family of statistical models indexed by a parameter vector $\paramVec$.
Different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population \citep{bansal07} which is simultaneously an avenue for powerful simplification and a source of model misspecification.
Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
In addition, elements of $\paramVec$ can either be chosen as constants, based on scientific reasoning and previous knowledge, or calibrated to observed data.
Suitable methodology for calibrating model parameters may depend on other modeling decisions.
While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:results}.

@@

\subsubsection{Model~3}
Model~3 is also of a probabilistic model.
In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
% We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest, and evidence for the former approach has been provided in previous studies \citep{king15}.
Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also good reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.
% Still, researchers that intend to use a mechanistic model to describe a dynamic system should design the model to be as realistic as their scientific understanding of the system and their computational abilities permit \eic{THIS SENTENCE SEEMS TO DISMISS THE ARGUMENTS IN THE PREVIOUS SENTENCE. PERHAPS THIS SENTENCE CAN JUST BE DELETED}.

@@

We then fit the panel version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
% Even with the simplification of Eq.~\myeqref{eq:model3:approx}, a significant amount of computational effort is needed to obtain a proper model fit.
The reported parameters were obtained using the PanelPOMP version of Model~3, and simulations for the various vaccination campaigns (Sec.~\ref{sec:results}) were obtained using these same parameters in the SpatPOMP version of the model.
% Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.

@@

\begin{figure}[ht]
<<Model3_PanelSims_Figure, echo=FALSE, fig.height=3>>=
ggplot() +
  geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
  geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
              alpha = 0.4) +
  facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
             labeller = dep_labeller) +
  ylab('Reported cholera cases') +
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

@
\caption{\label{h3panelsims}
Simulations from initial conditions using panelPOMP version of Model~3.
The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
}
\end{figure}

@@

@@

In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.

The use of a common benchmark may also be beneficial when developing models with varying spatial scale, as direct comparisons between models fit to data with different levels of spatial aggregation are meaningless (e.g. comparing Model~1 to Model~3).
\eic{I DON'T QUITE UNDERSTAND THIS YET. BENCHMARKS TEND TO APPLY TO SPECIFIC DATA SETS, SO CANNOT READILY BE USED TO COMPARE DIFFERENT LEVELS OF AGGREGATION. LIKELIHOOD BENCHMARKS CAN BE USED TO STUDY THE VALUE OF EXPLICIT SPATIAL COUPLING. HOW TO BENCHMARK ACROSS AGGREGATION LEVELS IS, I THINK, SOMETHING OF AN OPEN PROBLEM.}
Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.

@@

\begin{tabular}{|c|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3 \\ (panelPOMP)} & \thead{Model~3 \\ (SpatPOMP)}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  \multirow{2}{*}{$\Sexpr{myround(mod3_ll[1], 1)}$\footnotemark[2]} &
  $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  & ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[3]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ & $\Sexpr{as.character(mod2_n_params)}$ & \multirow{2}{*}{$\Sexpr{as.character(mod3_n_params)}$} & $\Sexpr{mod3_spatPomp_n_params}$ \\
 Fit Parameters & (\Sexpr{as.character(mod1_lee_n_params)}) & ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) && ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ & $\Sexpr{myround(mod2_aic, 1)}$ & \multirow{2}{*}{$\Sexpr{myround(mod3_aic, 1)}$\footnotemark[2]} & $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] & ($\Sexpr{myround(mod2_lee_aic, 1)}$) &  & ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[3]
\\
\hline
\thead{Log-ARMA(2,1) \\ Log-likelihood} &
  $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
\\
\hline
\end{tabular}
\caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.
\eic{I THINK THAT THE NOTES SHOULD GO IN THE CAPTION RATHER THAN BEING FOOTNOTES.}
\eic{WE SHOULD REVIEW THIS TABLE IN A MEETING. RIGHT NOW, I'M SURPRISED BY A FEW THINGS. WHY IS THE LEE ET AL MODEL 2 LIKELIHOOD HIGHER THAN OURS? IS THE (29) IN THE PANELPOMP COLUMN A TYPO, SINCE WE DON'T COMPARE TO LEE ET AL IN THIS COLUMN? NESTING RELATIONSHIP FOR THE SPATPOMP VERSION OF MODEL 3 MAKES IT HARD TO EXPLAIN THE RATHER LARGE DECREASE IN LOGLIK. AT FACE VALUE, THIS SEEMS TO MEAN THAT THE COUPLING IS SUFFICIENTLY POORLY SPECIFIED THAT WE ARE BETTER OFF WITHOUT IT. I SEEM TO RECALL THAT OUR UNDERSTANDING INVOLVES THE WAY PANELPOMP USES A COVARIATE THAT DEPENDS ON CASES IN OTHER DEPARTMENTS. SINCE LEE ET AL ALSO DO THAT, COULD IT BE THAT THEIR ANALYSIS IS MORE CLOSELY RELATED TO THE PANELPOMP VERSION THAN THE SPATPOMP VERSION?}
}
\end{table}

\jwc{NEW PARAGRAPH: }
Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).\footnotetext[1]{The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime.}
\footnotetext[2]{Parameters to department-specific models not provided by \citet{lee20}, as department fits were only used as an intermediary step to obtain parameter estimates of the coupled model.}
\footnotetext[3]{Model 3 was originally fit to only a subset of the data starting from March 2014 and did not include a large portion of data from Ouest in 2015-2016.
On this subset, the parameters provided by \cite{lee20} achieved a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
On this same subset of data, our model achieved a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.}

@@

\jwc{Replaced Ed's comment with this: }
We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
This perhaps demonstrates that while deterministic models describe the systems in a less realistic and useful way, the relative ease in fitting these models potentially results in fewer modeling errors.
As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.

\section{Results}\label{sec:results}

\jwc{SHOULD THESE FIRST TWO PARAGRAPHS BE MOVED TO THE DISCUSSION?}
\eic{YES. ALSO, THERE ARE RESULTS IN SEC.~3. PERHAPS, THE RAINFALL ANALYSIS IN SEC 4 COULD BE ANOTHER ASPECT OF THE SEC 3 ANALYSIS: COMPARISON OF MODEL FIT WITH OTHER LINES OF SCIENTIFIC EVIDENCE. THEN THE CURRENT SEC 4 COULD BE DISBANDED.
SOMETIMES THE METHODS-RESULTS-DISCUSSION FORMAT DOES NOT FIT SO NATURALLY WITH APPLIED STATS.
WE CAN DISCUSS THIS. ONE IDEA IS TO HAVE A SECTION NAMED SOMETHING LIKE "LESSONS LEARNED FROM MODELING THE HAITI EPIDEMIC" OR "ADVICE FOR FUTURE STUDIES". THE LATTER CAN SEEM PRESUMPTUOUS, AND SHOULD BE DONE WITH DUE CAUTION, BUT FROM ANOTHER PERSPECTIVE IF WE ARE NOT PROVIDING CONCRETE GUIDANCE FOR FUTURE WORK THEN WHAT ARE WE DOING? OUR RESPONSIBILITY IS TO MAKE THE STRONGEST CLAIM WE THINK WE CAN PROPERLY SUPPORT; THEN THE REFEREES CAN HAVE A GO AT PULLING THE CLAIM DOWN.}
\eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
}

@@

@@

% \eic{TODO: CHECK PARAMETERIZATION AND UNITS FOR $\obsOverdispersion$ AND $\sigmaProc$ TO MAKE THEM COMPARABLE}

@@

@@

The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models.
This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models for policy recommendations.
We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system in question and may improve the accuracy of model-based projections.
% TODO: "Improved model fits", or "Improved model fitting", based on Anna's suggestion.
Caution is nonetheless needed when making policy based on modeling conclusions, as model misspecification may invalidate conclusions.
% TODO: Citation above?

@@

We acknowledge there are limitations to this study;
one example was the inability to fit model parameters to the fully-coupled SpatPOMP version of Model~3.
Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to fit the SpatPOMP version of Model~3 in future work.

% \arc{Right now, the manuscript's main point/claim is a bit muddled (at least to me). I am not sure how to directly address this, but having a additional paragraph or two in the discussion may help the impact of this article/"drive our points home". I will think more on this in the next week, but maybe we could further emphasize the balance/dependence between policy, prediction, inference, model quality, and model fitting? Well informed policy depends upon reliable prediction which depends upon model quality/performance which depends upon rigorous evaluation and estimation, regardless of correct model specification (though that is also important/of great interest).}
% \jwc{Thanks, I agree that the main point isn't clear, and Ed has suggested as much also. I think that adding additional paragraphs would help this, but we are already 2 pages over the recommended submission length, so re-wording the text may be a better option than adding more to it... I'll also have to think about this more this week.}
% \eic{HERE'S A SUGGESTION FOR A CONCLUDING PARAGRAPH. I THINK IT IS SUPPORTED BY THE RESULTS...}
Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems. This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models. Care must be taken to ensure that the computationally intensive numerical calculations are carried out adequately. Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data. Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation. Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.

>>>>

# Check if RUN_LEVEL == 3 results exists, if so, use them.
# Include a flag to check if RUN_LEVEL == 3. If so, we might want to recompute.
# TODO: Only save parameters and likelihoods, not entire MIF objects.
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_fit.rds")) {
  lee1_epi_mif <- readRDS("model1/run_level_3/lee1_epi_fit.rds")
} else {
  lee1_epi_mif <- bake(
    file = paste0("model1/", rl_dir, "lee1_epi_fit.rds"),
    {
      foreach(
        i = 1:nrow(epi_params),
        .packages = c('pomp'),
        .combine = c
      ) %dopar% {
        r_params <- unlist(epi_params[i, ])
        coef(h1_epi) <- r_params
        mif2(
          h1_epi,
          Np = NP_MIF,
          Nmif = NMIF,
          cooling.fraction.50 = 0.5,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = epi_rw
        )
      }
    }
  )
}

@@

if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_epi_evals.rds")) {
  lee_epi_ll <- readRDS("model1/run_level_3/lee1_epi_evals.rds")
} else {
  lee_epi_ll <- bake(
    file = paste0("model1/", rl_dir, "lee1_epi_evals.rds"), {

@@

      lee_epi_ll
    }
  )
}

@@

# TODO: Only save parameter values and "likelihoods".
if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_fit.rds")) {
  lee1_end_mif <- readRDS("model1/run_level_3/lee1_end_fit.rds")
} else {
  lee1_end_mif <- bake(
    file = paste0("model1/", rl_dir, "lee1_end_fit.rds"),
    {
      foreach(
        i = 1:nrow(end_start),
        .packages = c('pomp'),
        .combine = c
      ) %dopar% {

@@

        coef(h1_end) <- r_params
        mif2(
          h1_end,
          Np = NP_MIF,
          Nmif = NMIF,
          cooling.type = "hyperbolic",
          cooling.fraction.50 = 0.05,  # Cooling set so that we will reach smaller rw.sd at the end than the end of the unit3 search.
          rw.sd = end_rw
        )
        # output[['m2']] <- m2
        #
        # coef(h1_end) <- coef(m2)
        # sims <- simulate(  # Step 5
        #   h1_end, nsim = N_LEE_SIMS, format = 'data.frame',
        #   params = coef(m2)
        # )
        # sims$param_set <- i
        #
        # output[['sims']] <- sims
        # output
      }
    }
  )
}

@@

if (RUN_LEVEL != 3 & file.exists("model1/run_level_3/lee1_end_evals.rds")) {
  lee_end_ll <- readRDS("model1/run_level_3/lee1_end_evals.rds")
} else {
  lee_end_ll <- bake(
    file = paste0("model1/", rl_dir, "lee1_end_evals.rds"), {

@@

      lee_end_ll
    }
  )
}

@@

We proceed by introducing Models~1--3 in Sec.~\ref{sec:models};
in Sec.~\ref{sec:methods}, we present a methodological approach to examining and refining these models, and then use improved model fits to project cholera incidence in Haiti under various vaccination scenarios. 
We then conclude with recommendations on the use of mechanistic models to inform policy decisions in Sec.~\ref{sec:advice}.

@@

Proposed mechanistic structures form a family of statistical models indexed by a parameter vector $\paramVec$.
Different values of $\paramVec$ can result in qualitative differences in the predicted behavior of the system.
The complex nature of biological systems necessitates a search for modeling assumptions that combine insightful simplicity with fidelity to biological reality.
For example, many models commonly used in epidemiology are motivated by reasoning about a homogeneous mixing population \citep{bansal07} which is simultaneously an avenue for powerful simplification and a source of model misspecification.
Other common considerations include whether the proposed model should be stochastic or deterministic; whether the model should have change points in parameter values or should otherwise make adjustments for changes through time in the dynamic system; and whether the proposed model should include any spatial heterogeneity at a scale permissible by the observed data.
In addition, elements of $\paramVec$ can either be chosen as constants, based on scientific reasoning and previous knowledge, or calibrated to observed data.
Suitable methodology for calibrating model parameters may depend on other modeling decisions.
% While this section is focused on the elements of $\paramVec$ that are calibrated to data, we note that the decision of which elements to fix and which to estimate has consequences for model interpretability, as discussed in Sec.~\ref{sec:science}.

@@

\subsubsection{Model~3}
Model~3 is also of a probabilistic model.
In this model, both the latent and observable processes can be factored into department specific processes which interact with each other.
% We denote the latent and measurement processes for Model~3 as $\bm{X}^{(3)}(t_{0:N}) = X^{(3)}_{1:U, 0:N}$, and $\bm{Y}^{(3)}(t_{1:N}) = Y_{1:U, 1:N}$ \eic{IS THIS NEEDED HERE?}.
The decision to model the system via metapopulation models versus a model aggregated to a larger spatial scale is one of great scientific interest.
Evidence for the former approach has been provided in previous studies \citep{king15}, including the specific case of heterogeneity between Haitian departments in cholera transmission \citep{collins14}.
Note that this evidence alone does not automatically discredit conclusions drawn via nationally aggregated models, as there are also good reasons to prefer a simple model over a complex one \citep{saltelli20,green15}.

@@

We then fit the spatially decoupled version of Model~3 using a slight modification of the PIF algorithm, which we call the Block Panel Iterated Filter (BPIF).
The BPIF algorithm is implemented in the \code{panelPomp} package \citep{breto20panelPomp} in the R programming language, and can be used by adding the argument \code{block = TRUE} in the \code{mif2} function.
Pseudo-code for this algorithm is provided in Algorithm~S1 in the supplementary material.
While the reported parameters were obtained using the decoupled representation of Model~3, simulations for the various vaccination campaigns were obtained using fully coupled version of the model.
% Simulations from the fitted version of the PanelPOMP model are shown in Fig.~\ref{h3panelsims}.

@@

% \begin{figure}[ht]
% ggplot() +
%   geom_line(data = h3_panel_quantiles, aes(x = date, y = q50 + 1), col = 'blue') +
%   geom_line(data = dep_plot_df, aes(x = date, y = cases + 1), col = 'black') +
%   geom_ribbon(data = h3_panel_quantiles, aes(x = date, ymin = q05 + 1, ymax = q95 + 1),
%               alpha = 0.4) +
%   facet_wrap(~factor(dep, levels = plot_order), nrow = 2,
%              labeller = dep_labeller) +
%   ylab('Reported cholera cases') +
%   theme(axis.title.x = element_blank()) +
%   scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2011-01-01"), as.Date("2019-01-01"), by = '2 years')) +
%   scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
%
% \caption{\label{h3panelsims}
% Simulations from initial conditions using panelPOMP version of Model~3.
% The black curve represents true case count, the blue curve the median of 500 simulations from the model, and the grey ribbons representing $95\%$ confidence interval.
% }
% \end{figure}

@@

In some cases, a possible benchmark model could be a generally accepted mechanistic model, but often no such model is available.
Because of this, we use a log-linear Gaussian ARMA model as an associative benchmark, as recommended by \citet{he10}.
The theory and practice of ARMA models is well developed, but the exponential growth and decay characteristic of biological dynamics suggests applying these linear models on a log scale.
Likelihoods of Models~1--3 and their respective ARMA benchmark models are provided in Table~\ref{tab:likes}.

@@

The use of benchmarks may also be beneficial when developing models with varying spatial scale, where a direct comparison between models likelihoods is meaningless.
In such a case, a benchmark model could be fit to each spatial resolution being considered, and each model compared to their respective benchmark.
Large advantages (shortcomings) in model likelihood relative to the benchmark for a given spatial scale that are not present in other spatial scales may provide weak evidence for (against) the spatial resolution under consideration.
While these comparisons don't provide a formal statistical test of the amount of spatial resolution needed for a given model, they can be quite informative.

@@

\begin{tabular}{|c|c|c|c|}
\hline
 & \thead{Model~1} & \thead{Model~2} & \thead{Model~3}
\\
\hline
\hline
\multirow{2}{*}{Log-likelihood} &
  $\Sexpr{myround(mod1_ll, 1)}$ &
  $\Sexpr{myround(mod2_ll, 1)}$ &
  $\Sexpr{myround(mod3_spatPomp_res$ll, 1)}$ \\
    & ($\Sexpr{myround(mod1_lee_ll, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(h2_fit$leeFit_ll, 1)}$) &
  ($\Sexpr{myround(max(lee3_results[["all_ll_evals"]]), 1)}$)\footnotemark[2]
\\
\hline
Number of &
  $\Sexpr{as.character(mod1_n_params)}$ &
  $\Sexpr{as.character(mod2_n_params)}$ &
  $\Sexpr{mod3_spatPomp_n_params}$ \\
 Fit Parameters &
 (\Sexpr{as.character(mod1_lee_n_params)}) &
 ($\Sexpr{as.character(h2_fit$leeFit_n_params)}$) &
 ($\Sexpr{as.character(lee3_results$n_params, 1)}$)
\\
\hline
\multirow{2}{*}{AIC} &
  $\Sexpr{myround(mod1_aic, 1)}$ &
  $\Sexpr{myround(mod2_aic, 1)}$ &
  $\Sexpr{myround(mod3_spatPomp_aic, 1)}$ \\
  & ($\Sexpr{myround(mod1_lee_aic, 1)}$)\footnotemark[1] &
  ($\Sexpr{myround(mod2_lee_aic, 1)}$) &
  ($\Sexpr{myround(mod3_lee_coupled_aic, 1)}$)\footnotemark[2]
\\
\hline
\thead{Log-ARMA(2,1) \\ Log-likelihood} &
  $\Sexpr{myround(ARMA_benchmarks$m1, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m2, 1)}$ &
  $\Sexpr{myround(ARMA_benchmarks$m3, 1)}$
\\
\hline
\end{tabular}
\caption{\label{tab:likes}Log-likelihood values for each models compared to their ARMA benchmarks.
Values in parenthesis are corresponding values using \citet{lee20} parameter estimates.  $\textsuperscript{1}$The reported likelihood is an upper bound of the likelihood of the \citet{lee20} model as it is the largest likelihood obtained using their parameter calibration regime. $\textsuperscript{2}$\cite{lee20} fit Model~3 to only a subset of the data (March 2014 onward) and did not include data from Ouest in 2015-2016.
On this subset, the \cite{lee20} fit has a likelihood of $\Sexpr{myround(max(lee3_results$sub_ll_evals), 1)}$.
On this same subset, our model has a likelihood of $\Sexpr{myround(mod3_spatPomp_res$subset_ll, 1)}$.
}
\end{table}

Similar to comparing log-likelihoods across models, an additional powerful diagnosis tool is the comparison of conditional log-likelihoods.
Conditional likelihoods, defined as the density $f_{Y_k | Y_1, \ldots, Y_{k - 1}}\big(Y_k = y_k^* | y_{1:k-1}^*\big)$, provide a basic description of how well the proposed model can describe each data point, given the previous observations.
Comparing these results across models---including benchmark models---can help researchers identify potential model deficiencies, or errors in the observed data.
Additional tools for assessing the goodness-of-fit of a model include plotting the effective sample size of each observation \jwc{REF???} and comparing any statistic of the observed data to simulations from the model, which is sometimes referred to as diagnostic probes (for example, the autocorrelation function (ACF), as was done in \cite{king15}).

@@

We note---the credit of deterministic models---that of the four fitted models in \cite{lee20}, Model~2 provides the most apparently accurate forecasts.
This perhaps demonstrates that while deterministic models describe the systems in a less realistic and useful way, the relative ease in fitting these models potentially results in fewer modeling errors.
As the data analysis becomes more refined, however, the deficiencies of deterministic models become increasingly apparent.

\subsection{Comparison of Model Fit to Oustide Scientific Evidence}\label{sec:science}

@@

\subsection{Predicting the Effect of Interventions}\label{sec:interventions}

@@

@@

\section{Advising policy via dynamic modeling}\label{sec:advice}

\eic{I THINK IT IS INTERESTING THAT THE MODEL-SUPPORTED SUGGESTION BY \citep{andrews11} HAS LARGELY PROVED ACCURATE (VACCINATION, NON-PHARMACEUTICAL INTERVENTIONS, AND INFRASTRUCTURE IMPROVEMENTS ALL HAVE A ROLE AND SHOULD BE FOLLOWED SIMULTANEOUSLY)DESPITE LEGITIMATE CONCERNS ABOUT WHETHER THE SCIENTIFIC SUPPORT FOR THE MODEL IS STRONG ENOUGH TO ROBUSTLY GUIDE POLICY \citep{grad12}.
THE MAIN CLAIM OF ANDREWS11 IS THAT FORECASTS AND POLICY NOT SUPPORTED BY A MODEL SEEM EVEN MORE UN-ROOTED THAN RECOMMENDATIONS BASED ON A WEAKLY SUPPORTED MODEL.
THE CONCLUSION FROM ALL THIS, WHICH I THINK IS SUPPORTED BY OUR RE-ANALYSIS OF LEE ET AL, IS THAT MECHANISTIC MODELS FITTED TO DATA ARE A STRONG WAY TO ENFORCE CONSISTENT THINKING, BUT WEAKER AS A WAY TO CHECK THAT YOU ARE ACTUALLY RIGHT DUE TO THE LIMITATIONS OF INFERRING CAUSAL MECHANISMS FROM OBSERVATIONAL DATA.
POLICY-MAKERS SHOULD FOLLOW AN APPROACH WHICH IS SELF-CONSISTENT AND CONSISTENT WITH PREVIOUS EVIDENCE THAT MIGHT WORK (WHICH MODELS CAN HELP WITH) BUT THEY MUST BE AGILE AS NEW EVIDENCE EMERGES.
MODELS, TOGETHER WITH DATA ANALYSIS, CAN ALSO HELP REVEAL WHEN THE EVIDENCE SUPPORTING A P0LICY IS BREAKING DOWN, FORCING A RE-THINK.
}

@@

The ongoing global COVID-19 pandemic has provided a clear example on how government policy may be affected by the conclusions of scientific models.
This article demonstrates that fitting appropriate scientific models remains a challenging statistical task, and therefore great care is needed when fitting scientific models for policy recommendations.
We provided a few suggestions that may aid the fitting of mechanistic models such as comparing model likelihoods to a benchmark.
Improved model fits allows for meaningful statistical inference that may provide valuable insight on a dynamic system and may improve the accuracy of model-based projections.
% TODO: "Improved model fits", or "Improved model fitting", based on Anna's suggestion.
Caution is nonetheless needed when making policy based on modeling conclusions, as model misspecification may invalidate conclusions.
% TODO: Citation above?

@@

We acknowledge there are limitations to this study;
one example was the inability to fit model parameters to the fully-coupled version of Model~3.
Promising theoretical and methodological developments \citep{ning21ibpf,ionides22} based on the Block Particle Filter \citep{rebeschini15} may potentially be used to obtain better parameter estimates for Model~3 in future work.

Inference for mechanistic time series models offers opportunities for understanding and controlling complex dynamic systems.
This case study has investigated issues requiring attention when applying powerful new statistical techniques that can enable statistically efficient inference for a general class of partially observed Markov process models. 
Care must be taken to ensure that the computationally intensive numerical calculations are carried out adequately. 
Once that is accomplished, care is required to assess what causal conclusions can properly be inferred given the possibility of alternative explanations consistent with the data.
Studies that combine model development with thoughtful data analysis, supported by a high standard of reproducibility, build knowledge about the system under investigation.
Cautionary warnings about the difficulties inherent in understanding complex systems \citep{saltelli20,ioannidis20} should motivate us to follow best practices in data analysis, rather than avoiding the challenge.
